---
title: "a.init_models"
author: "Dennis Kim"
date: "`r Sys.Date()`"
output: html_document
---

```{r libraries}
# call libraries 
library(here)
library(MASS)
library(nlme)
library(mgcv)
library(dplyr)
library(tidyr)
library(GGally)
library(ggplot2)
```

# 0. Data call
load the model fit data
```{r model fit}
# load data 
final_df <- readRDS(here::here("data/final_data", "Final_July_2025.rds"))

# factorize the cateogrical columns (animal_id, REALM, BIOME_NAME) 
final_df$BIOME_NAME <- as.factor(final_df$BIOME_NAME)
final_df$REALM <- as.factor(final_df$REALM)
final_df$animal_id <- as.factor(final_df$animal_id)

glimpse(final_df)
```

# 1. Model assumption checking 
```{r data1}
# data for model 1

# remove species that has small samples (4 species removed - now 13 species)
final_df1 <- final_df %>% dplyr::filter(! species %in% c("Cathartes burrovianus", "Gyps bengalensis", "Gyps rueppellii", "Sarcoramphus papa"))

# scaling NDVI and lifestcok
final_df1$wmLifestock_scale <- scale(final_df1$wmLifestock)

# grouping variables: 
## check the unique numbers per variables
unique(final_df1$BIOME_NAME) # 16 levels 
unique(final_df1$REALM) # 7 levels 
unique(final_df1$species)# 13 levels 
```

```{r colinearity}
# colinearity effects: no colinearity effects! 
corr_df <- final_df1 %>% dplyr::select(wmNDVI, wmLifestock, UDsizeKm2_99)

## plot colinearity
ggpairs(corr_df)

## remove the coor data 
rm(corr_df)
```

data for model 2: remove all the NAs in REALM group
```{r data2}
final_df2 <- final_df1 %>% filter(!is.na(REALM) & REALM != "N/A") %>% droplevels()

summary(final_df2)

# final data
final_df_clean <- na.omit(final_df2)
summary(final_df_clean)
```

# 2. Species levels
BAM - generalized addictive models for very large datasets: 
- Model UD size (km²) as a function of NDVI and Livestock
- Allow species-specific slopes and intercepts for NDVI and Livestock
- Account for spatial autocorrelation (from weighted centroids of UDs)
- Control for individual-level random effects (animal ID nested in species)
- Not focus on temporal trends, but monthly grouping is available
- Eventually, visualize species-wise trends over NDVI and Livestock

```{r species level}
# 1. GAM with species level
model1 <- bam(
  log10(UDsizeKm2_99) ~
    
    # Species-specific smooths over NDVI and Livestock (different slopes + intercepts)
    s(wmNDVI, by = species, k = 5) +
    s(wmLifestock_scale, by = species, k = 5) +
    
    # Include species as a factor to allow different intercepts
    species +
    
    # Spatial smooth (varying by species)
    s(UDwMeanLongitude, UDwMeanLatitude, by = species, k = 100) +
    
    # Random effect for individual animals
    s(animal_id, bs = "re"),
    
  data = final_df_clean,
  family = gaussian(),
  method = "fREML", # faster and stable for BAM
  discrete = TRUE   # good for large datasets
)

# save the model 1
saveRDS(model1, file = "model1.rds")
```

```{r}
# 2. GAM with REALM 
model2 <- bam(
  log10(UDsizeKm2_99) ~
    
    # Environmental Predictors with **Species-Specific** Smooths
    s(wmNDVI, by = species, k = 5) +
    s(wmLifestock_scale, by = species, k = 5) +
    
    # Environmental Predictors with **REALM-Specific** Smooths
    s(wmNDVI, by = REALM, k = 5)+
    s(wmLifestock_scale, by = REALM, k =5)+
    
    # Allows different intercepts per species and REALM (fixed effect)
    species + REALM +
    
    # Spatial smooth (varying by species)
    s(UDwMeanLongitude, UDwMeanLatitude, by = species, k = 100) +
    
    # Random effect for individual animals
    s(animal_id, bs = "re"),
    
  data = final_df_clean,
  family = gaussian(),
  method = "fREML", # faster and stable for BAM
  discrete = TRUE   # good for large datasets
)

# save the model 2
saveRDS(model2, file = "model2.rds")
```


